manuscript_outline

### 2025-06-27 docs/manuscript/paper.Rmd
add references, copy edits

Morgan M. Sparks^1^, Brian C. Leavell^2^, Bryan M. Maitland^1^

@@

^2^ Brian's deets here

@@

Development in poikilothermic organisms is strongly temperature-dependent, particularly during early life stages prior to exogenous feeding. This thermally constrained physiology produces a near-mechanistic relationship between ambient temperature and developmental rate, allowing for predictive models of phenology. However, traditional models---often developed under constant laboratory temperatures---struggle to account for the variable thermal regimes experienced by wild populations. To address this, [@sparks2019] introduced the effective value model, a framework that integrates daily temperature data to predict developmental timing under fluctuating conditions. Building on this foundation, the R package **hatchR** was developed to operationalize the effective value model and enable its widespread application, initially focusing on fishes. In this paper, we expand the scope of **hatchR**, demonstrating its applicability across a broad range of poikilothermic taxa, including amphibians, reptiles, and aquatic invertebrates. We also illustrate how this framework supports both applied management goals---such as predicting hatching windows for conservation and monitoring---and basic research on ecological and evolutionary questions, including phenological plasticity, thermal adaptation, and life history strategies. By providing a generalizable, open-source tool, **hatchR** bridges the gap between physiological theory and real-world application, offering a robust platform for predicting developmental phenology in diverse and changing environments.

@@

Historically, models of developmental phenology in wild populations have relied on the accumulation of degree days (thermal sums) above a species-specific threshold. While widely used, degree-day models are fundamentally linear and fail to account for the non-linear nature of development, leading to biased predictions under fluctuating temperature regimes [@neuheimer2007]. This is particularly problematic for long-developing species or for organisms in variable climates, where embryos incubating at colder temperatures often require fewer thermal units to complete development than those incubating in warmer environments. Moreover, traditional models often assume temporally invariant developmental rates and do not incorporate diel temperature variability, which can significantly affect development in systems with large day--night thermal swings **[CITATION needed, perhaps\*\* <https://doi.org/10.1186/1472-6785-13-18> \*\*...though I wonder why this is brought up, because hatchR doesn't appear to address DTV?, since it auto-sums to daily averages? or am I misunderstanding? It seems like DTV might be better brought up in the discussion as a caveat?]**. Finally, developmental reaction norms may vary across populations due to local adaptation or phenotypic plasticity, further complicating predictions [@sparks2017; @angilletta2009].

To address these limitations in fishes, [@sparks2019] introduced the effective value model---an approach that captures the non-linear, temperature-dependent nature of development by integrating experimental data into a daily developmental rate function. Building on that work, the R package **hatchR** [CITE hatchR] was developed to operationalize the effective value framework and make it accessible to researchers and managers. **hatchR** uses temperature-performance curves generated from lab-based thermal rearing trials to estimate the proportion of development achieved each day under observed temperature conditions. These daily contributions are summed until a cumulative threshold (typically 1.0) is reached, marking the completion of a developmental stage. The package includes both an R-based workflow and a user-friendly Shiny web interface to facilitate broad adoption across research and management communities.

While hatchR was originally developed for coldwater fishes, the approach is broadly applicable to any poikilothermic organism whose development conforms to a monotonic or unimodal temperature--rate relationship, including many amphibians, reptiles, and invertebrates. In this manuscript, we expand the taxonomic scope of **hatchR**, compiling thermal developmental data from the literature and demonstrating how effective value modeling can be applied to diverse taxa. We also present three case studies illustrating how **hatchR** can inform ecological, evolutionary, and management questions---ranging from understanding thermal adaptation and phenological mismatches to forecasting hatch timing in conservation planning. These applications are increasingly relevant as freshwater ecosystems face intensifying and interacting stressors, including climate change, habitat alteration, and invasive species, all of which can disrupt developmental timing and population dynamics [@craig2017; ADD ANOTHER BROAD PAPER; **also, seems unnecessary to limit to freshwater ecosystems?**].

@@

Effective value models estimate developmental progress based on empirically derived relationships between temperature and development rate. These relationships are typically parameterized by raising organisms at constant temperatures in laboratory settings, then fitting a non-linear function that captures the temperature-dependence of development. This function can then be reciprocated to produce daily "effective values"---units of development---that accumulate over time.

@@

```{r tail-frog, fig.width=8, fig.height=3, fig.align='center', fig.cap="Custom hatching phenology model for coastal tailed frogs (Ascaphus truei). Panel A represents the model fit and raw data used to generate the effective value model and panel B are the effective values for daily temperatures between 6 and 20 °C. A dashed line with a 0.01 increase for every degree increase is included for reference."}
# Brown 1975, Comparitive Biochemistry and Physicology Part A: Physiology
# https://www.sciencedirect.com/science/article/pii/030096297590033X

@@

Developmental phenology can be predicted in two ways using hatchR. The `predict_phenology()` function estimates the date at which development completes, given a known reproductive event (e.g., oviposition or spawning; `spawn.date`) and daily temperature data. Alternatively, the `predict_spawn()` function works in reverse---estimating the likely date of a reproductive event based on a known or observed hatch/emergence date. This is especially useful for retrospective analyses or for estimating breeding windows from field observations.

@@

The effective value modeling framework is broadly applicable across poikilothermic taxa and can be used to address a wide array of ecological, evolutionary, and management questions. While this manuscript focuses on three illustrative case studies, many additional opportunities exist. Below, we provide a non-exhaustive compilation of species for which thermal developmental data are available in the peer-reviewed or grey literature. These sources can be used to parameterize custom models via the `fit_model()` function in **hatchR**, allowing users to build tailored effective value models for diverse taxa. The examples span four taxonomic classes---amphibians, reptiles, insects, and crustaceans---and demonstrate the breadth of systems to which this approach can be applied. While the case studies we present here are intended as proof-of-concept, we anticipate that applications will continue to expand alongside the growing availability of temperature--development datasets.

**BCL NOTE: Thoughts on adding this marine copepod paper?** <https://doi.org/10.1086/662174>

@@

## Ecological Appication: Spatial and Temporal Trends in Development

@@

**BCL NOTE: It feels like this section jumps into the methods a bit too quick, without providing context. At the start of this first paragraph, it would be worth providing a broad, 2-3 sentence overview of the "what" and "why" of this specific ecological application... e.g. (rough idea), "Examining variation in development across space and time is critical to understand how organisms respond to ecological factors such as climate change [CITATION]. Here, we first demonstrate an application of hatchR to predict hatching times from a spatiotemporally extensive dataset. Using these predictions, we then show how one can examine putative drivers of variation in development and discuss their application in forecasting future responses."**

@@

The dataset included `r nrow(model_data)` observations of predicted days to hatching, drawn from `r length(unique(model_data$comid))` stream reaches (labeled with Common Identifiers, i.e., COMIDs) nested within four mountain ranges: Olympic National Park, the Cascade Mountains, the Siskiyou Mountains, and the Willapa Hills. Each COMID had between 20 and 30 years of predictions.

**BCL NOTES: 1) It isn't clear to me what the original data consists of and what was modeled. It sounds like the original data is just oviposition observation dates, effective values, and stream temps. But in the paragraph above, "The dataset included observations of predicted days to hatching" makes me question what was data and what was modeled. It seems like the model generated predicted days to hatching?, which would not be "observations" and would not be part of the "dataset". (I think it is the "observations of predicted days" that is throwing me, as one doesn't observe predictions... I mean, technical you could, but I don't think that's what is meant.) But below, it states that the "model explained X% of the variation in developmental timing", suggesting that there was original hatching data to which the predicted hatching times were compared? Clearly, I'm confused. lol. Could just be me, but maybe worth being more specific and hand-holdy. 2) I explained "COMID" above, as not everyone will know what that means (especially if anticipating a broad readership). But please change if there's a better way.**

@@

**BCL (minor) NOTE: *Ascaphus truei* needs italics in Fig. 2, but not sure how to edit**

@@

p.01 <- means_all |> 
  ggplot(aes(year, Mean, ymin = CI_low, ymax = CI_high)) +
  geom_ribbon(alpha = 0.05, color = NA) +
  geom_line(size = 1.1)  +
  labs(x = "Year", y = "Mean of Days to Hatch") +
  theme_classic()

@@

Our results indicate developmental rates were highly variable over different physical habitats (streams originating in lowland hills compared with those originating in mountain ranges with permanent snow fields), with sites in Olympic National Park consistently having the longest development times and the greatest interannual variability (SD \~ 11 days), while lowland streams in the Willapa Hills developed fastest with minimal variability (SD \< 3 days). These results demonstrate the capacity of **hatchR** to detect spatiotemporal phenological variation and forecast responses to shifting climate patterns.

@@

We applied the custom parameterized models to two hypothetical spring temperature regimes with mean daily temperatures of 16 °C (cool) and 24 °C (warm). Under the cool regime, wood frogs hatched earliest while green frogs lagged behind by up to eight days---nearly double the developmental duration (Figure \@ref(fig:litho-pheno)A). However, under the warm regime, predicted hatch dates clustered tightly, with all species hatching within one day of each other (Figure \@ref(fig:litho-pheno)B). The effective value slopes (Figure \@ref(fig:litho-pheno)C) reveal these underlying differences in developmental response curves, serving as species-specific linearized reaction norms. Notably, the rank order of developmental timing reverses across the two thermal regimes, a hallmark of strong G×E interaction.

@@

This example demonstrates how **hatchR** can be used to explore evolutionary and ecological differences in thermal sensitivity across species. When combined with phylogenetic or trait-based analyses, such reaction norms can reveal how developmental strategies align with species' ecological niches, helping to explain the persistence of closely related species across thermal gradients.

@@

Here, we used data from a common garden experiment on cabbage beetles (*Colaphellus bowringi*) reared at six constant temperatures (16--28 °C), spanning populations collected along a \~3,440 km latitudinal gradient in China [@tang2017]. Although all beetles were reared under identical conditions, developmental rates differed among populations, particularly at cooler temperatures. Populations from southern latitudes---where temperatures are generally warmer---developed more quickly than northern populations, a pattern consistent with cogradient variation, where genetic and environmental effects on a phenotype act in the same direction [@conover2009; @sparks2022].

To visualize these patterns, we fit custom effective value models for five of the six populations (excluding Longnan, which duplicated the values from Xiushui). The resulting linearized effective value curves (Figure \@ref(fig:beetles)) show that beetles from Xiushui County (southernmost) had consistently faster development rates across temperatures than their more northerly counterparts. These differences are especially pronounced at cooler temperatures, suggesting that local adaptation may only manifest under certain environmental conditions.

@@

We highlight the extensibility of **hatchR** through a compilation of parameterizable species across four taxonomic classes and through three case studies that illustrate ecological, evolutionary, and conservation applications. These case studies span field data, phylogenetic comparisons, and experimental designs, demonstrating the tool's flexibility across data types and biological questions.

As with any model of developmental timing, it is important to recognize that temperature is not the sole determinant of phenological events. For embryos, environmentally cued hatching is seen across taxa, wherein variation in hatch timing and developmental stage at hatching can result from myriad abiotic and biotic triggers [Warkentin 2011 <https://doi.org/10.1093/icb/icr017>]. While **hatchR** focuses specifically on temperature-dependent development, it can be integrated with complementary datasets or experimental manipulations to investigate these more complex, context-dependent interactions.

We also encourage users to carefully consider how models are parameterized. As discussed in greater detail in Sparks et al. (2025), we recommend using experimental data with at least four distinct temperature treatments spanning a wide thermal range. This is critical for capturing the nonlinear shape of developmental rate curves and avoiding overfitting or extrapolation errors. Many studies we encountered---particularly older or gray literature---relied on only two temperature treatments (e.g., [@qualls1998; @kozák2009]), which we do not recommend for model fitting due to the inability to detect curvature.

Given the wide potential applicability of **hatchR**, we especially emphasize its utility for rapidly developing species. For example, as shown in our phylogenetic case study, frog species with short incubation periods can be reared across multiple treatments in a single month, making them ideal for both research and teaching applications. Researchers can use these models to investigate thermal adaptation, phenological mismatch, or range-wide trait variation. In educational settings, developmental models could be built for *Daphnia*, insects, or amphibians to introduce students to reaction norms, thermal performance, and evolutionary trade-offs. Moreover, while **hatchR** currently focuses on pre-feeding stages, the effective value approach could potentially be extended to post-hatch phenological events---particularly when energy input is standardized (e.g., individuals reared under *ad libitum* conditions) [@lillehammer1986].

@@

Although initially developed for fishes, **hatchR** is a broadly applicable tool for modeling temperature-dependent development in poikilotherms. It addresses key limitations of traditional degree-day approaches by accounting for non-linear thermal response curves and enabling predictions under fluctuating thermal conditions. As global temperatures rise and the phenology of organisms shifts in response, tools like **hatchR** will be increasingly valuable for anticipating the ecological and evolutionary consequences of climate change. While we provide numerous species-level parameterizations and demonstrate a set of representative case studies, the full range of possible applications---across taxa, ecosystems, and educational settings---extends far beyond what we present here.

>>>>

Morgan M. Sparks^1,3^, Brian Z. Leavell^2^, Bryan M. Maitland^1^

@@

^2^Brian's deets here

^3^Corresponding author, Morgan.Sparks\@usda.gov

keywords: hatchR, developmental phenology, local adaptation, reaction norm, R, Shiny

@@

Development in poikilothermic organisms is strongly temperature-dependent, particularly during early life stages prior to exogenous feeding. This thermally constrained physiology produces a near-mechanistic relationship between ambient temperature and developmental rate, allowing for predictive models of phenology. However, traditional models—often developed under constant laboratory temperatures—struggle to account for the variable thermal regimes experienced by wild populations. To address this, [@sparks2019] introduced the effective value model, a framework that integrates daily temperature data to predict developmental timing under fluctuating conditions. Building on this foundation, the R package **hatchR** was developed for fishes to operationalize the effective value model and enable its widespread application. In this paper, we expand the scope of **hatchR**, demonstrating its applicability across a broad range of poikilothermic taxa, including amphibians, reptiles, and aquatic invertebrates. We also illustrate how this framework supports both applied management goals—such as predicting hatching windows for conservation and monitoring—and basic research on ecological and evolutionary questions, including phenological plasticity, thermal adaptation, and life history strategies. By providing a generalizable, open-source tool, **hatchR** bridges the gap between physiological theory and real-world application, offering a robust platform for predicting developmental phenology in diverse and changing environments.

@@

Historically, models of developmental phenology in wild populations have relied on the accumulation of degree days (thermal sums) above a species-specific threshold. While widely used, degree-day models are fundamentally linear and fail to account for the non-linear nature of development, leading to biased predictions under fluctuating temperature regimes [@neuheimer2007; @steel2012]. This is particularly problematic for long-developing species or for organisms in variable climates, where embryos incubating at colder temperatures often require fewer thermal units to complete development than those incubating in warmer environments. Moreover, traditional models often assume temporally invariant developmental rates and do not incorporate diel temperature variability, which can significantly affect development in systems with large day–night thermal swings. Finally, developmental reaction norms may vary across populations due to local adaptation or phenotypic plasticity, further complicating predictions [@sparks2017; @angilletta2009].

To address these limitations in fishes, [@sparks2019] introduced the effective value model—an approach that captures the non-linear, temperature-dependent nature of development by integrating experimental data into a daily developmental rate function. Building on that work, the R package **hatchR** [@sparks2025] was developed to operationalize the effective value framework and make it accessible to researchers and managers. **hatchR** uses temperature-performance curves generated from lab-based thermal rearing trials to estimate the proportion of development achieved each day under observed temperature conditions. These daily contributions are summed until a cumulative threshold is reached (typically 1.0), marking the completion of a developmental stage. The package includes both an R-based workflow and a user-friendly Shiny web interface to facilitate broad adoption across research and management communities.

While hatchR was originally developed for coldwater fishes, the approach is broadly applicable to any poikilothermic organism whose development conforms to a monotonic or unimodal temperature–rate relationship, including many amphibians, reptiles, and invertebrates. In this manuscript, we expand the taxonomic scope of **hatchR**, compiling thermal developmental data from the literature and demonstrating how effective value modeling can be applied to diverse taxa. We also present three case studies illustrating how **hatchR** can inform ecological, evolutionary, and management questions—ranging from understanding thermal adaptation and phenological mismatches to forecasting hatch timing in conservation planning. These applications are increasingly relevant as freshwater ecosystems face intensifying and interacting stressors, including variable climates, habitat alteration, and invasive species, all of which can disrupt developmental timing and population dynamics [@craig2017; ADD ANOTHER BROAD PAPER].

@@

Effective value models estimate developmental progress based on empirically derived relationships between temperature and development rate. These relationships are typically parameterized by raising organisms at constant temperatures in laboratory settings, then fitting a non-linear function that captures the temperature-dependence of development. This function can then be reciprocated to produce daily “effective values”—units of development—that accumulate over time.

@@

```{r tail-frog, fig.width=8, fig.height=3, fig.align='center', fig.cap="Custom hatching phenology model for coastal tailed frogs (Ascaphus truei).Panel A represents the model fit and raw data used to generate the effective value model and panel B are the effective values for daily temperatures between 6 and 20 °C. A dashed line with a 0.01 increase for every degree increase in included for reference."}
# Brown 1975, Comparitive Biochemistry and Physicology Part A: Physiology
# https://www.sciencedirect.com/science/article/pii/030096297590033X

@@

Developmental phenology can be predicted in two ways using hatchR. The `predict_phenology()` function estimates the date at which development completes, given a known reproductive event (e.g., oviposition or spawning; `spawn.date`) and daily temperature data. Alternatively, the `predict_spawn()` function works in reverse—estimating the likely date of a reproductive event based on a known or observed hatch/emergence date. This is especially useful for retrospective analyses or for estimating breeding windows from field observations.

@@

The effective value modeling framework is broadly applicable across poikilothermic taxa and can be used to address a wide array of ecological, evolutionary, and management questions. While this manuscript focuses on three illustrative case studies, many additional opportunities exist. Below, we provide a non-exhaustive compilation of species for which thermal developmental data are available in the peer-reviewed or grey literature. These sources can be used to parameterize custom models via the `fit_model()` function in **hatchR**, allowing users to build tailored effective value models for diverse taxa. The examples span four taxonomic classes—amphibians, reptiles, insects, and crustaceans—and demonstrate the breadth of systems to which this approach can be applied. While the case studies we present here are intended as proof-of-concept, we anticipate that applications will continue to expand alongside the growing availability of temperature–development datasets.

@@

## Ecological Application: Spatial and Temporal Trends in Development

@@

@@

The dataset included `r nrow(model_data)` observations of predicted days to hatching, drawn from `r length(unique(model_data$comid))` stream reaches (COMIDs) nested within four mountain ranges: Olympic National Park, the Cascade Mountains, the Siskiyou Mountains, and the Willapa Hills. Each COMID had between 20 and 30 years of predictions.

@@

@@

p.01 <- means_all |> 
  ggplot(aes(year, Mean, ymin = CI_low, ymax = CI_high)) +
  geom_ribbon(alpha = 0.05, color = NA) +
  geom_line(size = 1.1)  +
  lims(y = c(20,27)) +
  labs(x = "Year", y = "Mean of Days to Hatch") +
  theme_classic()

@@

Our results indicate developmental rates were highly variable over different physical habitats (streams originating in lowland hills compared with those originating in mountain ranges with permanent snow fields), with sites in Olympic National Park consistently having the longest development times and the greatest interannual variability (SD \~ 11 days), while lowland streams in the Willapa Hills developed fastest with minimal variability (SD \< 3 days). These results demonstrate the capacity of **hatchR** to detect spatiotemporal phenological variation and forecast potential shifts under shifting climate patterns.

@@

We applied the custom parameterized models to two hypothetical spring temperature regimes with mean daily temperatures of 16 °C (cool) and 24 °C (warm). Under the cool regime, wood frogs hatched earliest while green frogs lagged behind by up to eight days—nearly double the developmental duration (Figure \@ref(fig:litho-pheno)A). However, under the warm regime, predicted hatch dates clustered tightly, with all species hatching within one day of each other (Figure \@ref(fig:litho-pheno)B). The effective value slopes (Figure \@ref(fig:litho-pheno)C) reveal these underlying differences in developmental response curves, serving as species-specific linearized reaction norms. Notably, the rank order of developmental timing reverses across the two thermal regimes, a hallmark of strong G×E interaction.

@@

This example demonstrates how **hatchR** can be used to explore evolutionary and ecological differences in thermal sensitivity across species. When combined with phylogenetic or trait-based analyses, such reaction norms can reveal how developmental strategies align with species’ ecological niches, helping to explain the persistence of closely related species across thermal gradients.

@@

Here, we used data from a common garden experiment on cabbage beetles (*Colaphellus bowringi*) reared at six constant temperatures (16–28 °C), spanning populations collected along a \~3,440 km latitudinal gradient in China [@tang2017]. Although all beetles were reared under identical conditions, developmental rates differed among populations, particularly at cooler temperatures. Populations from southern latitudes—where temperatures are generally warmer—developed more quickly than northern populations, a pattern consistent with cogradient variation, where genetic and environmental effects on a phenotype act in the same direction [@conover2009; @sparks2022].

To visualize these patterns, we fit custom effective value models for five of the six populations (excluding Longnan, which duplicated the values from Xiushui). The resulting linearized effective value curves (Figure \@ref(fig:beetles)) show that beetles from Xiushui County (southernmost) had consistently faster development rates across temperatures than their more northerly counterparts. These differences are especially pronounced at cooler temperatures, suggesting that local adaptation may only manifest under certain environmental conditions. Moreover, the consistent rank order of developmental rates across the populations indicates heritable differences among populations which have and can continue to be selected upon in changing environments.

@@

We highlight the extensibility of **hatchR** through a compilation of parameterizable species across four taxonomic classes and through three case studies that illustrate ecological, evolutionary, and conservation applications. These case studies span field data, phylogenetic comparisons, and experimental designs, demonstrating the tool’s flexibility across data types and biological questions.

As with any model of developmental timing, it is important to recognize that temperature is not the sole determinant of phenological events. Other environmental cues—such as photoperiod, substrate quality, hydrology, salinity, or exposure to contaminants (e.g., metals or endocrine-disrupting chemicals)—can modulate developmental rates or induce thresholds not captured in thermal response curves (NEED REF). Biotic cues may also play a role; for example, the presence of predators can delay hatching in some amphibians as an adaptive strategy [*e.g.*, @sih2004]. While **hatchR** focuses specifically on temperature-dependent development, it can be integrated with complementary datasets or experimental manipulations to investigate these more complex, context-dependent interactions.

We also encourage users to carefully consider how models are parameterized. As discussed in greater detail in documentation for @sparks2025, we recommend using experimental data with at least four distinct temperature treatments spanning a wide thermal range. This is critical for capturing the nonlinear shape of developmental rate curves and avoiding overfitting or extrapolation errors. Many studies we encountered—particularly older or gray literature—relied on only two temperature treatments (e.g., [@qualls1998; @kozák2009]), which we do not recommend for model fitting due to the inability to detect curvature.

Given the wide potential applicability of **hatchR**, we especially emphasize its utility for rapidly developing species. For example, as shown in our phylogenetic case study, frog species with short incubation periods can be reared across multiple treatments in a single month, making them ideal for both research and teaching applications. Researchers can use these models to investigate thermal adaptation, phenological mismatch, or range-wide trait variation. In educational settings, developmental models could be built for *Daphnia*, insects, or amphibians to introduce students to reaction norms, thermal performance, and evolutionary trade-offs. Moreover, while **hatchR** currently focuses on pre-feeding stages, the effective value approach could potentially be extended to post-hatch phenological events—particularly when energy input is standardized (*e.g.*, individuals reared under *ad libitum* conditions) [@lillehammer1986].

@@

Although initially developed for fishes, **hatchR** is a broadly applicable tool for modeling temperature-dependent development in poikilotherms. It addresses key limitations of traditional degree-day approaches by accounting for non-linear thermal response curves and enabling predictions under fluctuating thermal conditions. As global temperatures rise and the phenology of organisms shifts in response, tools like **hatchR** will be increasingly valuable for anticipating the ecological and evolutionary consequences of climate change. While we provide numerous species-level parameterizations and demonstrate a set of representative case studies, the full range of possible applications—across taxa, ecosystems, and educational settings—extends far beyond what we present here.

# Acknowledgements

We thank N. Karraker, D. Pilliod, M. Hayes, R.B. Bury, and B. Norman for their help in translating sites for the tailed frog example to locations that could be aligned with COMIDs. The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of USFS . Any mention of trade names, products, or services does not imply an endorsement by the U.S. government or USFS. USFS does not endorse any commercial products, services or enterprises.
